JD notes on effects

### 2022-07-08 manuscript/body.tex
GEtting on airplane

% Use "Eq" instead of "Equation" for equation citations. 
%% BC: use \EREF{} for equations and \FREF{} for figures
% Best to use a command that combines the name and the reference
\section{Introduction}

@@

Representative values for a focal predictor are generally chosen using quantiles, or equally spaced values, for a continuous predictor, or an exhaustive set of levels for a categorical predictor.
The central estimates are then calculated by holding the non-focal predictors at their reference point (values chosen for the non-focal predictors, e.g., means) while varying the focal predictor, with the goal that the estimates represent how the model responds to the changes in the focal predictor \citep{fox2009effect, hanmer2013behind}. This procedure generates -- \emph{predictor effects} \citep{fox2009effect}, \emph{marginal predictions} \citep{leeper2017package} or \emph{estimated marginal means} \citep{lenth2018package}. In this article, we refer to these quantities as the \emph{central estimates} for an ``average case'' in a population; the associated confidence intervals will either describe a prediction or effect plot. In the effect plots, we can also choose an anchor (value chosen for the focal predictor).

For linear models (including models with complex interactions), the averaging is done on the linear scale, i.e., linear averaging. As a result, the \emph{bias} in the expected (mean) predictions can easily be corrected using \emph{mean-based reference point}, i.e., averaging of the non-focal model variables -- currently not implemented in commonly used \proglang{R} software packages. The mean of the central estimates is the prediction at the \emph{model center}. In this case, the estimates are usually consistent with the observed values. However, in models with non-linear link functions, this is not always the case. In particular, when dealing with GL(M)Ms with non-linear link functions, the averaging is done on the non-linear scale, i.e., non-linear averaging, making generating correct predictions much harder. This is the bias in the expected mean prediction induced by the non-linear transformation of the response variable. One way to address this is to make predictions on the linear predictor scale and back-transform to the original scale. However, the back-transformation may either result in biased predictions or requires some approximation, for example, second-order Taylor approximation implemented in \pkg{emmeans} \citep{lenth2018package}. An alternative to the mean-based reference point is the \emph{whole-sample-based} approach, discussed in detail later, which involves computing the prediction over the population of non-focal predictors and then averaging across the values of the focal predictor \citep{hanmer2013behind}.

@@

\jd{I should edit if we can finally consense on nomenclature. I feel it's my fault.}
\begin{itemize}
\item \textbf{Input variables}: Refers to the observed (or scientific) variables underlying an inference or exploration. For example, the regression models described by \EREF{simple_inter_higher_no_interaction} and \EREF{simple_inter_higher} both have $3$ input variables -- $x_1, x_2, x_3$.
\item \textbf{Focal and non-focal predictors}: The focal predictor refers to the input variable we are interested in. For multivariate models, all other variables other than focal predictors are non-focal predictors. For example, consider model described by \EREF{simple_inter_higher_no_interaction}. Suppose we are interested in the effect of $x_1$ on the predicted $y$, then $x_1$ is the focal predictor while $x_2$ and $x_3$ are non-focal predictors.
\item \textbf{Model matrix}: Refers to the design matrix whose rows include all combination of input variables. Consider an example for three hypothetical households -- the first household head is a Christian with an income of $\$ 50$, the second household head is Muslim with an income of $\$ 100$ while the third household head is a Jew with an income of $\$ 77$. Suppose we want to model the household size (\code{hhsize}) as function of these household characteristics, i.e., $$\mathrm{hhsize} = \beta_0 + \beta_1\times\mathrm{income} + \beta_{2[r]}\times\mathrm{religion} + \mathrm{error}.$$ The model matrix corresponding to this model is given by
$$\begin{bmatrix}{}
 Intercept & income & religionJew & religionMuslim \\
 1 & 50 & 0 & 0 \\
  1 & 100 & 0 & 1 \\
  1 & 77 & 1 & 0 \\
\end{bmatrix}.$$ The first column represents the constant term in our model, $\beta_0$. For continuous input variables, the representation in the model matrix is the same as the corresponding input variables (for example second column representing income). For categorical variables, however, by default, the model matrix  creates additional dummy variables using the reference cell parameterization. This means that, if an input variable has $L$ factor levels, then there will be $L-1$ dummy columns representing all but the first level created in the model matrix. In our example, the column \code{religion} is missing and instead we have \code{religionJew} and \code{religionMuslim}; the missing category \code{religionChristian} is treated as the reference category.

@@

\item \textbf{Model center:} A point corresponding to a column-wise mean of the model matrix (the mean of one or more model variables). Also referred to as center point. We refer to the prediction approach that uses model center as mean-based. The center point for a set of model variables corresponding to an input variable may not represent a possible value of the input variable -- the case in \FREF{justify_plots}B. In simple linear models (without) interaction, the mean of the input variable can be used as a proxy for the center point. If there are more than one model variables associated with the focal input variable, however, the model center does not correspond to any given value of the focal predictor, and we do not necessarily get the ``nice'' crossing we get in simpler cases, see \FREF{pred_cubic_plots}.
\item \textbf{Reference point:} Corresponds to the value or values chosen for non-focal predictors, when estimating the predictions and effects. Typically the center point (mean-based), but can instead be a whole-sample of quantiles or observations. Appropriately choosing the reference point is very crucial in generating correct predictions (for bias correction) especially when dealing with non-linear link functions.
\item \textbf{Anchor:} The value chosen for the focal predictor when estimating effect confidence intervals. The anchor choice does not affect the central estimates. Typically chosen as the center point of the model variables corresponding to the focal predictor but other sensible values can be chosen too. If we are trying to focus on effects, the anchor from which we calculate confidence intervals becomes important. This is not true for the predictions. We can choose any anchor, but the center point is a natural and stable choice.
\item \textbf{Marginal effects:} In simple linear models with no interaction terms, the coefficient estimates are simple and directly interpretable as the expected change in outcome for a unit change in focal input variable. This is the \emph{unconditional marginal effect} and it is constant across all the observations and levels of all other input variables. Consider models described by \EREF{simple_inter_higher_no_interaction} and \EREF{simple_inter_higher}. The marginal effect of $x_2$ in model~\ref{eq:simple_inter_higher_no_interaction} is $\frac{\partial y}{\partial x_2} = \beta_2$. On the other hand, the marginal effect of $x_2$ in model~\ref{eq:simple_inter_higher} is given by $\frac{\partial y}{\partial x_2} = \beta_2 + \beta_{23}x_3$. In other words, if there are no interactions, the marginal effect of $x_2$ on $y$ is constant, while, if there are interactions in the model, the marginal effect of a change in $x_2$ on $y$ depends on the value of the other \emph{conditioning} predictor, $x_3$. In essence, computation of marginal effects requires the use of partial derivatives and can be considered as the slope of a multi-dimensional surface with respect to the surface of interest \cite{leeper2017interpreting}. 
\end{itemize}

>>>>

% Use "Eq" instead of "Equation" for equation citations. 
%% BC: use \EREF{} for equations and \FREF{} for figures
%% JD: Why do you capitalize these? It's scary.
% Best to use a command that combines the name and the reference
\section{Introduction}

@@

Representative values for a focal predictor are generally chosen using quantiles, or equally spaced values, for a continuous predictor; or an exhaustive set of levels for a categorical predictor.
The central estimates are then calculated by holding the non-focal predictors at a reference point while varying the focal predictor, with the goal that the estimates represent how the model responds to the changes in the focal predictor \citep{fox2009effect, hanmer2013behind}. These values have been called: \emph{predictor effects} \citep{fox2009effect}, \emph{marginal predictions} \citep{leeper2017package} or \emph{estimated marginal means} \citep{lenth2018package} \jd{Is this also the same as least-squares means?}. In this article, we refer to these quantities as the \emph{central estimates} of the outcome.

\jd{It looks like you're trying to introduce column averaging and non-linear averaging in the same \P. How can we simplify?}
For linear models (including models with complex interactions), the averaging is done on the linear scale, i.e., linear averaging. As a result, the \emph{bias} in the expected (mean) predictions can easily be corrected using \emph{mean-based reference point}, i.e., averaging of the non-focal model variables -- currently not implemented in commonly used \proglang{R} software packages. The mean of the central estimates is the prediction at the \emph{model center}. In this case, the estimates are usually consistent with the observed values. However, in models with non-linear link functions, this is not always the case. In particular, when dealing with GL(M)Ms with non-linear link functions, the averaging is done on the non-linear scale, i.e., non-linear averaging, making generating correct predictions much harder. This is the bias in the expected mean prediction induced by the non-linear transformation of the response variable. One way to address this is to make predictions on the linear predictor scale and back-transform to the original scale. However, the back-transformation may either result in biased predictions or requires some approximation, for example, second-order Taylor approximation implemented in \pkg{emmeans} \citep{lenth2018package}. An alternative to the mean-based reference point is the \emph{whole-sample-based} approach, discussed in detail later, which involves computing the prediction over the population of non-focal predictors and then averaging across the values of the focal predictor \citep{hanmer2013behind}.

@@

\begin{itemize}
\item \textbf{Input variables}: Refers to the observed (or scientific) variables underlying an inference or exploration. For example, the regression models described by \EREF{simple_inter_higher_no_interaction} and \EREF{simple_inter_higher} both have $3$ input variables -- $x_1, x_2, x_3$.
\item \textbf{Focal predictors}: We call the input variable on the x-axis of an outcome plot the focal predictor. 
Any other predictor variables are “non-focal” predictors. 
For example, consider the model described by \EREF{simple_inter_higher_no_interaction}. 
If we are focusing on the effect of $x_1$ on the response $y$, then $x_1$ is the focal predictor while $x_2$ and $x_3$ are non-focal predictors. \jd{Is this trivial example necessary? Is any example necessary here?}

\item \textbf{Model matrix}: Refers to the design matrix whose rows include all combination of input variables. Consider an example for three hypothetical households -- the first household head is a Christian with an income of $\$ 50$, the second household head is Muslim with an income of $\$ 100$ while the third household head is a Jew with an income of $\$ 77$. Suppose we want to model the household size (\code{hhsize}) as function of these household characteristics, i.e., $$\mathrm{hhsize} = \beta_0 + \beta_1\times\mathrm{income} + \beta_{2[r]}\times\mathrm{religion} + \mathrm{error}.$$ The model matrix corresponding to this model is given by
$$\begin{bmatrix}{}
 Intercept & income & religionJew & religionMuslim \\
 1 & 50 & 0 & 0 \\
  1 & 100 & 0 & 1 \\
  1 & 77 & 1 & 0 \\
\end{bmatrix}.$$ The first column represents the constant term in our model, $\beta_0$. For continuous input variables, the representation in the model matrix is the same as the corresponding input variables (for example second column representing income). For categorical variables, however, by default, the model matrix  creates additional dummy variables using the reference cell parameterization. This means that, if an input variable has $L$ factor levels, then there will be $L-1$ dummy columns representing all but the first level created in the model matrix. In our example, the column \code{religion} is missing and instead we have \code{religionJew} and \code{religionMuslim}; the missing category \code{religionChristian} is treated as the reference category.

@@

\item \textbf{Model center:} A point corresponding to a column-wise mean of the model matrix (the mean of one or more model variables). Also referred to as center point. We refer to the prediction approach that uses model center as mean-based. The center point for a set of model variables corresponding to an input variable may not represent a possible value of the input variable -- the case in \FREF{justify_plots}B. In simple linear models with no interactions, the mean of the input variable can be used as a proxy for the center point \jd{Is it a “proxy” or is it the thing??}. If there are more than one model variables associated with the focal input variable, however, the model center does not correspond to any given value of the focal predictor \jd{This seems like bad repetition}, and we do not necessarily get the ``nice'' crossing we get in simpler cases, see \FREF{pred_cubic_plots}.
\item \textbf{Reference point:} Corresponds to the value or values chosen for non-focal predictors, when estimating the predictions and effects. Typically the center point (mean-based), but can instead be a whole-sample of quantiles or observations. Appropriately choosing the reference point is very crucial in generating correct predictions (for bias correction) especially when dealing with non-linear link functions.
\item \textbf{Anchor:} The value chosen for the focal predictor when estimating effect confidence intervals. The anchor choice does not affect the central estimates. Typically chosen as the center point of the model variables corresponding to the focal predictor but other sensible values can be chosen too. If we are trying to focus on effects, the anchor from which we calculate confidence intervals becomes important. This is not true for the predictions. We can choose any anchor, but the center point is a natural and stable choice.
\item \textbf{Marginal effects:} In simple linear models with no interaction terms, the coefficient estimates are simple and directly interpretable as the expected change in outcome for a unit change in focal input variable. This is the \emph{unconditional marginal effect} and it is constant across all the observations and levels of all other input variables. Consider models described by \EREF{simple_inter_higher_no_interaction} and \EREF{simple_inter_higher}. The marginal effect of $x_2$ in model~\ref{eq:simple_inter_higher_no_interaction} is $\frac{\partial y}{\partial x_2} = \beta_2$. On the other hand, the marginal effect of $x_2$ in model~\ref{eq:simple_inter_higher} is given by $\frac{\partial y}{\partial x_2} = \beta_2 + \beta_{23}x_3$. In other words, if there are no interactions, the marginal effect of $x_2$ on $y$ is constant, while, if there are interactions in the model, the marginal effect of a change in $x_2$ on $y$ depends on the value of the other \emph{conditioning} predictor, $x_3$. In essence, computation of marginal effects requires the use of partial derivatives and can be considered as the slope of a multi-dimensional surface with respect to the surface of interest \cite{leeper2017interpreting}. 
\end{itemize}
